---
title: "gapfraction: R functions for airborne laser scanning data of forests"
author:
  name: "Adam Erickson, PhD Candidate, University of British Columbia"
  email: "adam.erickson@ubc.ca"
date: "Updated: May 28, 2017"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{gapfraction: R functions for airborne laser scanning data of forests}
  %\VignetteEncoding{UTF-8}
---

> "The science of today is the technology of tomorrow."
[Edward Teller]("http://news.stanford.edu/news/2003/september24/tellerobit-924.html")

The **gapfraction** package for **R** is designed for processing airborne laser scanning (ALS) light-detection-and-ranging (LiDAR) data of forests. The package was produced for a chapter of my doctoral thesis at University of British Columbia^[Erickson, A. (2017) A hybrid modeling approach to simulating past-century understory solar irradiation in Alberta, Canada. University of British Columbia.]. The package is designed to be used with LiDAR data pre-processed with **LAStools**^[LAStools: http://rapidlasso.com/lastools/], **USDA Fusion**^[USDA Fusion: http://forsys.cfr.washington.edu/fusion/fusionlatest.html], or the new **lidR** package from Jean-Romain Roussel^[lidR: https://github.com/Jean-Romain/lidR]. The main input to functions are LAS or LAZ format height-normalized point clouds, typically LiDAR plots corresponding to field plots. The functions are designed to accept either paths to height-normalized LAS/LAZ files or `data.frame`/`data.table` objects created with **lidR** or **rLiDAR** from Carlos Silva^[rLiDAR: https://cran.r-project.org/web/packages/rLiDAR/index.html] packages. Functions automatically detect whether a path or object is input for the `las` parameter.

## Functions included

The **gapfraction** package implements a new barycentric interpolation spike-free canopy height model (CHM) algorithm based on Khosravipour et al. (2013)^[Khosravipour et al. (2013) Development of an algorithm to generate a LiDAR pit-free canopy height model. http://www.riegl.com/uploads/tx_pxpriegldownloads/khosravipour_SilviLaser2013.pdf] and (2016)^[Khosravipour et al. (2016) Generating spike-free digital surface models using LiDAR raw point clouds: A new approach for forestry applications. http://www.sciencedirect.com/science/article/pii/S0303243416300873], in addition to two new LiDAR metrics of canopy gap fraction ($P_o$), several individual tree crown (ITC) detection algorithms, canopy distance and direction metrics, effective leaf area index ($L_e$) and apparent clumping index ($ACI$) estimation methods. Four mathematical fisheye (hemispherical) lens models are implemented: equi-angular, equi-distant, stereographic, and orthographic. Finally, a new `radial.grid` function is included for polar plots, while a modified sun path plotting function is also provided. An alphabetical list of functions contained in the **gapfraction** package is provided below.

- `chm` - Simple canopy height model
- `chm.sf` - Barycentric interpolation spike-free canopy height model
- `dd.canopy` - Euclidean distance and direction to nearest canopy pixel from plot center
- `dd.crown` - Euclidean distance and direction to nearest tree crown from plot centers
- `itc.mw` - Variable-window individual tree crown detection
- `itc.mw.h` - Hierarchical variable-window individual tree crown detection
- `itc.wat` - Watershed segmentation individual tree crown detection
- `itc.wat.h` - Hierarchical watershed segmentation individual tree crown detection
- `Le.bl` - Ground-to-total-return ratio with a spherical leaf angle distribution
- `Le.n` - Contact frequency and vertical canopy cover-based effective LAI
- `radial.grid.hemi` - Modified radial.grid function supporting hemispherical lens geometries
- `P.hv` - Hemipsherical Voronoi canopy gap fraction
- `P.hv.par` - Parallel hemispherical Voronoi canopy gap fraction with SOCKS
- `P.pdn` - Point-density-normalized canopy gap fraction, effective LAI (Le), and ACI
- `sun.path` - Modified solar position plots of Thomas Steiner
- `vcc.aci` - Above-height cover index of vertical canopy cover
- `vcc.bl` - Beer-Lambert-Law-modified intensity-return ratio of vertical canopy cover
- `vcc.cv` - 2-D Cartesian Voronoi vertical canopy cover
- `vcc.fci` - First-echo cover index of vertical canopy cover
- `vcc.fr` - Canopy-to-first-return ration of vertical canopy cover
- `vcc.ir` - Intesity-return ratio of vertical canopy cover
- `vcc.p` - Canopy-to-total-pixel ratio of vertical canopy cover
- `vcc.r` - Canopy-to-total-return ratio of vertical canopy cover
- `vcc.sci` - Solberg's cover index of vertical canopy cover

## Canopy light transmission equations implemented

####Point-density normalized gap fraction:
$$\begin{align*}
P_\text{pdn} &= \sum_{i=1}^{n} \sum_{j=1}^{n}
\left(\frac{{n_\text{first return}}_{i,j}} {\rho_\text{first return} / {{A_\text{sector}}_{i,j}}} \times \frac{A_{\text{sector}_{i,j}}}{A_\text{hemisphere}}\right)
\end{align*}$$

####Hemispherical Voronoi gap fraction above a canopy height threshold with an equidistant lens projection:
$$\begin{align*}
x_\text{equidistant} &= \left( \frac{x}{\sqrt{x^2+y^2}} \right)
\times atan2 \left( \sqrt{x^2+y^2},z \right) \\
y_\text{equidistant} &= \left( \frac{y}{\sqrt{x^2+y^2}} \right)
\times atan2 \left( \sqrt{x^2+y^2},z \right) \\[4pt]
p &= \left( x_\text{equidistant}, y_\text{equidistant} \right) \\[4pt]
V_a(A_i) &= p|d_a (p,A_i) \le d_a(p,A_j), j \ne i, j=1, \cdots, n \\
P_\text{hv} &= 1 - \sum_{i=1}^{n} V_a(A_i)
\text{ where } h_i \ge h_\text{threshold}
\end{align*}$$

####Beer-Lambert Law canopy gap fraction:
$$\begin{align*}
L_e &= 2\sum_{i=1}^n - \ln\overline{P(\theta_i)}\cos{\theta}
\frac{\sin\theta_i}{\sum_{j=1}^n\sin\theta_j} \\
P_o &= \sum_{i=1}^n \exp \left (\frac{-L_eG(\theta_i)}{\cos\theta_i} \right)
\end{align*}$$

####Canopy-to-total-return ratio:
$$\begin{align*}
VCC_r &= \frac {\sum_{i=1}^n n_{\text{all}_i} \gt 1.25 \text{ m}}
{\sum_{i=1}^n n_{\text{last}_i} + n_{\text{single}_i}}
\end{align*}$$

####Canopy-to-total-first-return ratio:
$$\begin{align*}
VCC_\text{fr} &= \frac {\sum_{i=1}^n n_{\text{all}_i} \gt 1.25 \text{ m}}
{\sum_{i=1}^n n_{\text{first}_i}}
\end{align*}$$

####Intensity-return ratio:
$$\begin{align*}
VCC_\text{ir} &= \frac {\sum_{i=1}^n I_{\text{ground}_i} \gt 1.25 \text{ m}}
{\sum_{i=1}^n I_{\text{all}_i}}
\end{align*}$$

####Beer's Law modified intensity-return ratio:
$$\begin{align*}
VCC_\text{bl} &= \frac {\frac
{\sum_{i=1}^n I_{\text{ground single}_i} }
{\sum_{i=1}^n I_{\text{all}_i}} + \sqrt \frac
{\sum_{i=1}^n I_{\text{ground last}_i}}
{\sum_{i=1}^n I_{\text{all}_i}} }
{\frac
{\sum_{i=1}^n I_{\text{first}_i} + I_{\text{single}_i}}
{\sum_{i=1}^n I_{\text{all}_i}} + \sqrt \frac
{\sum_{i=1}^n I_{\text{intermediate}_i} + I_{\text{last}_i}}
{\sum_{i=1}^n I_{\text{all}_i}} }
\end{align*}$$

####Above-height cover index:
$$\begin{align*}
VCC_\text{aci} &= \frac
{\sum_{i=1}^n n_{\text{single}_i} + n_{\text{all}_i \gt 1.25 \text{ m}} +
n_{\text{intermediate}_i} + n_{\text{last}_i}}
{\sum_{i=1}^n n_{\text{all}_i}}
\end{align*}$$

####First-echo cover index:
$$\begin{align*}
VCC_\text{fci} &= \frac
{\sum_{i=1}^n n_{\text{single}_i \gt 1.25 \text{ m}} +
n_{\text{first}_i \gt 1.25 \text{ m}}}
{\sum_{i=1}^n n_{\text{single}_i} +
n_{\text{first}_i}}
\end{align*}$$

####Solberg's cover index:
$$\begin{align*}
VCC_\text{sci} &= \frac
{\sum_{i=1}^n n_{\text{single}_i \gt 1.25 \text{ m}} +
\frac{1}{2} (n_{\text{first}_i \gt 1.25 \text{ m}} +
n_{\text{last}_i \gt 1.25 \text{ m}})}
{\sum_{i=1}^n n_{\text{single}_i} + \frac{1}{2}
(n_{\text{first}_i} + n_{\text{last}_i})}
\end{align*}$$

####Canopy-to-total-pixel ratio:
$$\begin{align*}
VCC_p &= \frac
{\sum_{i=1}^n n_{\text{CHM}_i \gt 1.25 \text{ m}}}
{\sum_{i=1}^n n_{\text{CHM}_i}}
\end{align*}$$

####Cartesian-Voronoi fractional cover:
$$\begin{align*}
V_a(A_i) &= p|d_a (p,A_i) \le d_a(p,A_j), j \ne i, j=1, \cdots, n \\
P_\text{cv} &= 1 - \sum_{i=1}^{n} V_a(A_i)
\text{ where } h_i \ge h_\text{threshold}
\end{align*}$$

# Getting started

## Data pre-processing

By far, the simplest option for pre-processing data is to follow the **lidR** package [tutorial]("https://github.com/Jean-Romain/lidR/wiki/lasnormalize"), which utilizes **LAStools** under the hood. For LiDAR data without ground point classifications, height-normalized point clouds can be produced either with two **LAStools** command line functions, `lasground` and `lasheight`, or with three functions in **USDA Fusion**, `GroundFilter`, `GridSurfaceCreate`, and `CanopyModel`. If the ground points are already classified then you only need to use the `lasheight` function of **LAStools**, while the process for Fusion still requires the same three functions. Hence, I recommend that you use **LAStools**, as its ground point classification algorithm is also superior to that of **Fusion**, producing more accurate height-normalized point clouds. This is because **Fusion** uses the Kraus and Pfeifer (1998) algorithm^[Kraus and Pfeifer (1998) Determination of terrain models in wooded areas with airborne laser scanner data. http://www.sciencedirect.com/science/article/pii/S0924271698000094], while **LAStools** implements an optimized version of the Axelsson (1999) algorithm^[Axelsson (1999) Processing of laser scanner data—algorithms and applications. http://www.sciencedirect.com/science/article/pii/S0924271699000088]. For more information, read Maguya, Junttila, and Kauranne (2014)^[Maguya, Junttila, and Kauranne (2014) http://www.mdpi.com/2072-4292/6/7/6524]. An example application of `lasground` and `lasheight`, implemented in **Command Prompt** or **Bash**, is provided below.

```
lasground -i lidar.las -o lidar_g.las 
lasheight -i lidar_g.las -o lidar_n.las -replace_z 
```

In order to run these functions, you need to istall **LAStools**. For Windows, don't forget to add the **LAStools** `bin` directory to your system path^[http://www.computerhope.com/issues/ch000549.htm]. For a single LiDAR plot, this is simple to run without leaving your R session. You can call these functions using the `system` function included in base **R**, as shown below.

```
setwd('C:/lidar')
system(lasground -i lidar.las -o lidar_g.las)
system(lasheight -i lidar_g.las -o lidar_n.las -replace_z)
```

To loop through LAS files in a folder, the syntax follows this pseudocode:

```
folder <- 'C:/lidar'
files  <- list.files(folder, pattern="\\.las$", full.names=TRUE)

for (i in 1:length(files)) {
  file   <- files[i]
  basenm <- basename(file)
  filenm <- strsplit(basenm,'.',fixed=TRUE)[[1]][1]
  ground <- paste(folder,filenm,'_ground.las',sep='')
  htnorm <- paste(folder,filenm,'_norm.las',sep='')
  system(paste('lasground -i ',file,' -o ',ground, sep=''))
  while (!file.exists(ground)) { Sys.sleep(1) }
  system(paste('lasheight -i ',ground,' -o ',htnorm,' -replace_z', sep=''))
  while (!file.exists(htnorm)) { Sys.sleep(1) }
}
```
What this `for` loop does is read in each LAS file path, extract the name of the file without extension, create the filenames of the ground and height-normalized outputs, execute `lasground`, wait for the output, execute `lasheight` using the ground file as the input, wait for the output, then proceed to the next iteration. The code should be simple to parallelize using the `foreach` package, with each fork running in a new Command Prompt or Bash window.

Again, using the **lidR** package is much simpler!

## Example data

After loading the gapfraction package with the `library` function, the example data can be loaded by calling `data(las)`. The included data consists of fake $X,Y,Z$ coordinates in UTM 11N and meters, along with 8-bit unsigned interger values for intensity and return number. The data consists of one 100-meter diameter LiDAR plot. Based on previous research, I recommend using plots minimally of this size for comparison to ground data (e.g., hemispherical photography) to capture edge effects.

Let's get started!

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE)
```

```{r lastable, echo=FALSE, results='asis'}
require(gapfraction)
data(las)
knitr::kable(head(las,10))
```

Once the data is loaded, you can proceed to call functions from the `gapfraction` package.

## Example usage

#### Computing hemispherical Voronoi gap fraction

```{r Phv, echo=3:4, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6.5}
require(gapfraction)
data(las)
P.hv(las=las, model="equidist", thresh.val=1.25, thresh.var="height", reprojection=NA,
     pol.deg=5, azi.deg=45, col="height", plots=T, plots.each=T, plots.save=F)
```

#### Comparison of canopy height model (CHM) algorithms

```{r chmplots, echo=3:4, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6}
require(gapfraction)
data(las)
chm(las)
chm.sf(las)
```

#### Creating a pit-free CHM and performing individual tree crown (ITC) detection with the standard variable-window and watershed algorithms

```{r stditc, echo=4:5, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6}
require(gapfraction)
data(las)
chm  <- chm.sf(las, silent=TRUE)
mw   <- itc.mw(chm, ht2rad=function(x) 0.15746*x)
wat  <- itc.wat(chm, ht2rad=function(x) 0.15746*x)
```

#### Creating a stacked pit-free CHM and performing individual tree crown (ITC) detection with the hierarchical variable-window and watershed algorithms

```{r hieritc, echo=4:5, messages=FALSE, warnings=FALSE, verbose=FALSE}
require(gapfraction)
data(las)
chm  <- chm.sf(las, stacked=TRUE, silent=TRUE)
mw   <- itc.mw.h(chm, ht2rad=function(x) 0.15746*x, silent=TRUE)
wat  <- itc.wat.h(chm, ht2rad=function(x) 0.15746*x, silent=TRUE)
```

The `itc` algorithms store values for `trees` and `crown.area` in a two-element named numeric vector.

Happy laser scanning :)
