---
title: "gapfraction: R functions for LiDAR metrics of canopy light transmission"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=FALSE)
```

<!-- [Voronoi equation reference]("http://e-collection.library.ethz.ch/view/eth:3321") -->

<br>

> "The science of today is the technology of tomorrow."
[Edward Teller]("http://news.stanford.edu/news/2003/september24/tellerobit-924.html")

The **gapfraction** package for **R** is designed for processing airborne laser scanning (ALS) light-detection-and-ranging (LiDAR) data of forests. Many functions may work for terrestrial laser scanning data too. The package was produced for a chapter of Adam Erickson's doctoral thesis at University of British Columbia^[Erickson, A. (2017) A hybrid modeling approach to simulating past-century understory solar irradiation in Alberta, Canada. University of British Columbia.]. The package is designed to be used with LiDAR data pre-processed with **LAStools**^[LAStools: http://rapidlasso.com/lastools/], **USDA Fusion**^[USDA Fusion: http://forsys.cfr.washington.edu/fusion/fusionlatest.html], or the new **lidR** package from Jean-Romain Roussel^[lidR: https://github.com/Jean-Romain/lidR].

The main input to functions are LAS or LAZ format height-normalized point clouds, typically LiDAR plots corresponding to field plots. The functions are designed to accept either paths to height-normalized LAS/LAZ files or `data.frame`/`data.table` objects created with **lidR** or **rLiDAR** from Carlos Silva^[rLiDAR: https://cran.r-project.org/web/packages/rLiDAR/index.html] packages. Functions automatically detect whether a path or object is input for the `las` parameter. If using objects, the column order and naming conventions must follow those of **lidR** and **rLiDAR**.

<br>

# Getting started

<br>

## Installation

To install the **gapfraction** package, first ensure that you have a working development environment, as the package depends on the **lidr** package. Required software is listed below.

- Windows: Rtools.exe
- Mac: Xcode
- Linux: R development branch (e.g., r-devel or r-base-dev)

Next, install **gapfraction** from the R command line:

```
devtools::install_github("adam-erickson/gapfraction", dependencies=TRUE)
```

<br>

## Example data

After loading the gapfraction package with the `library` function, the example data can be loaded by calling `data(las)`. The included data consists of fake $X,Y,Z$ coordinates in UTM 11N and meters, along with 8-bit unsigned interger values for intensity and return number. The data consists of one 100-meter diameter LiDAR plot. Based on previous research, I recommend using plots minimally of this size for comparison to ground data (e.g., hemispherical photography) to capture edge effects.

Let's get started!

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width=6, fig.height=6, echo=FALSE, warning=FALSE, message=FALSE)
```

```{r lastable, echo=FALSE, results='asis'}
require(gapfraction)
data(las)
knitr::kable(head(las,10))
```

Once the data is loaded, you can proceed to call functions from the `gapfraction` package.

<br>

## Example usage

#### Computing hemispherical Voronoi gap fraction

<br>

```{r Phv, echo=3:4, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6.5}
require(gapfraction)
data(las)
P.hv(las=las, model="equidist", thresh.val=1.25, thresh.var="height", reprojection=NA,
     pol.deg=5, azi.deg=45, col="height", plots=T, plots.each=T, plots.save=F)
```

#### Comparison of canopy height model (CHM) algorithms

<br>

```{r chmplots, echo=3:4, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6}
require(gapfraction)
data(las)
chm(las)
chm.sf(las)
```

#### Creating a pit-free CHM and performing individual tree crown (ITC) detection with the standard variable-window and watershed algorithms

<br>

```{r stditc, echo=4:5, messages=FALSE, warnings=FALSE, verbose=FALSE, fig.show='hold', fig.width=6, fig.height=6}
require(gapfraction)
data(las)
chm  <- chm.sf(las, silent=TRUE)
mw   <- itc.mw(chm, ht2rad=function(x) 0.15746*x)
wat  <- itc.wat(chm, ht2rad=function(x) 0.15746*x)
```

#### Creating a stacked pit-free CHM and performing individual tree crown (ITC) detection with the hierarchical variable-window and watershed algorithms

<br>

```{r hieritc, echo=4:5, messages=FALSE, warnings=FALSE, verbose=FALSE}
require(gapfraction)
data(las)
chm  <- chm.sf(las, stacked=TRUE, silent=TRUE)
mw   <- itc.mw.h(chm, ht2rad=function(x) 0.15746*x, silent=TRUE)
wat  <- itc.wat.h(chm, ht2rad=function(x) 0.15746*x, silent=TRUE)
```

The `itc` algorithms store values for `trees` and `crown.area` in a two-element named numeric vector.

Please submit any issues and pull requests at the **gapfraction** page on GitHub at [adam-erickson/gapfraction]("https://github.com/adam-erickson/gapfraction").
